---
title: "Rscript DeSeq2 Analysis SN + cytokines"
output: html_document
date: "2023-03-23"
editor_options: 
  chunk_output_type: console
---

If necessary, download the required libraries
```{r}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(version = "3.20")

BiocManager::install("DESeq2")
BiocManager::install("edgeR")
BiocManager::install("tidyverse")
BiocManager::install("ggrepel")
install.packages("ggplot2")
install.packages("ggfortify")
BiocManager::install("RColorBrewer")
BiocManager::install("pheatmap")
install.packages("reshape")
BiocManager::install("PCAtools")
#BiocManager::install("apeglm")
#BiocManager::install("annotables")
install.packages("openxlsx")
BiocManager::install("EnhancedVolcano")
```

Load required libraries
```{r}
library("DESeq2")
library("edgeR")
library("tidyverse")
library("ggrepel")
library("ggplot2")
library("ggfortify")
library("RColorBrewer")
library("pheatmap")
library("reshape")
library(openxlsx)
library("EnhancedVolcano") #install BioC::Manager and then EnhancedVolcano
#library("annotables") #genesymbols aan ID matchen
```

Background info and help with DESeq2
```{r}
#browseVignettes(package = "DESeq2")
```

Load metadata and counts files
```{r}
# load the metadata and rowcount.csv file
setwd("G:/My Drive/RNAseq/Carli_SN+Cytokines/Analysis Carli")

metadata <- read.csv("Metadata.csv", row.names=1)
counts <- read.csv("Rawcounts.csv", row.names=1)

# Filter metadata to include only rows with treatment identifier "IL4IL13 or TGF"
metadata <- metadata %>% filter(Treatment %in% c("CTR", "IL4IL13"))
# Filter metadata to include only rows with Experiment identifier SN21-25
metadata <- metadata %>% filter(Experiment %in% c("B", "C", "D", "E", "F"))
# Save the filtered metadata
##write.csv(metadata, "Metadata_filtered_IL4IL13-noSN20.csv")

# Extract the row names (these will be the column names in rawcounts)
selected_columns <- rownames(metadata)
# Select the corresponding columns from rawcounts
counts <- counts %>% select(all_of(selected_columns))
# Save the filtered rawcounts data if needed
##write.csv(counts, "Rawcounts_filtered_IL4IL13-noSN20.csv")

# check if order of samples is the same in metadata and count table
rownames(metadata)
colnames(counts)
all(rownames(metadata) == colnames(counts))
```

Create DESEq2 object
```{r}
# remember that DESeq2 will output results for the last term in the formula 
ddsoverall <- DESeqDataSetFromMatrix(countData = counts, 
                                  colData = metadata, 
                                  design = ~ Experiment + Treatment)
ddsoverall$Treatment <- factor(ddsoverall$Treatment, levels = c("CTR", "IL4IL13"))
```

Prefiltering
```{r}
keep <- rowSums(counts(ddsoverall)) >= 10
ddsoverall <- ddsoverall[keep,]
```

DEA + fit raw counts to negative binomial model, count normalization 
```{r}
ddsoverall <- DESeq(ddsoverall)

# Generate data table of epithelial data
resoverall <- results(ddsoverall)
resultsNames(ddsoverall)

#Plot dispersion plot
plotDispEsts(ddsoverall)
```

Count normalization #waarom normaliseer je 2x? DESEQ2 doet dit al? - kan je niet de genormaliseerde data van Deseq2 niet exporteren?
```{r}
normalized_counts <- counts(ddsoverall, normalized = TRUE)
log2_normalized_counts <- log2(normalized_counts + 1)  # Adding 1 to avoid log(0)
##write.xlsx(as.data.frame(log2_normalized_counts), "log2normalized_counts_IL4IL13-noSN20.xlsx", rowNames = TRUE)
```

Log transformation 
```{r}
vsd_overall <- vst(ddsoverall, blind = TRUE)
```

Heatmap of sample correlation values
```{r}
vsd_mat <- assay(vsd_overall)
vsd_cor <- cor(vsd_mat)

pheatmap(vsd_cor, annotation = select(metadata, Treatment), clustering_distance_rows = "euclidean")
```

Plot a ScreePlot from the PCA analysis 
```{r}
# Perform PCA
pca <- prcomp(t(vsd_cor), scale. = TRUE)

# Calculate the proportion of variance explained by each component
explained_variance <- pca$sdev^2 / sum(pca$sdev^2)

# Create a data frame to store the results
pca_variance_df <- data.frame(PC = 1:length(explained_variance),
                              Variance = explained_variance)

# ggplot2 scree plot
ggplot(pca_variance_df, aes(x = PC, y = Variance)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Components", y = "Proportion of Variance Explained",
       title = "Scree Plot") +
  scale_x_continuous(breaks = pca_variance_df$PC)
```

PCA of PC2 vs PC3 - and - PC1 vs PC3
```{r}
# Extract the assay data from variance-stabilizing transformation (VST)
pca_data_vsd <- plotPCA(vsd_overall, intgroup = "Treatment", returnData = TRUE)

# pca_data_vsd now contains PC1 and PC2 (you can extract PC3 from vsd)
# Perform PCA manually on the VST-transformed data to get PC3
pca_manual <- prcomp(t(assay(vsd_overall)))

# Add PC3 to the PCA data returned by plotPCA
pca_data_vsd$PC3 <- pca_manual$x[, 3]
# Add PC4 to the PCA data returned by plotPCA
pca_data_vsd$PC4 <- pca_manual$x[, 4]
# Add PC5 to the PCA data returned by plotPCA
pca_data_vsd$PC5 <- pca_manual$x[, 5]

# Plot PC2 vs PC3 using ggplot2
ggplot(pca_data_vsd, aes(x = PC2, y = PC3, color = Treatment, label = name)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) + 
  theme_minimal() +
  labs(x = "PC2", y = "PC3", title = "PCA: PC2 vs PC3")

# Plot PC1 vs PC3 using ggplot2 - adapt to plot PC-values of choice
ggplot(pca_data_vsd, aes(x = PC1, y = PC3, color = Treatment, label = name)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) + 
  theme_minimal() +
  labs(x = "PC1", y = "PC3", title = "PCA: PC1 vs PC3")
```

PCA (different options) directly on the vst values instead 
```{r}
# simple PCA 
plotPCA(vsd_overall, intgroup = c("Experiment", "Treatment"))
plotPCA(vsd_overall, intgroup = "Treatment")
plotPCA(vsd_overall, intgroup = "Experiment")


plotPCA(vsd_overall, intgroup = c("Experiment", "Treatment"))
plotPCA(vsd_overall, intgroup = "Treatment")

# labeled PCA (can change text with repel for white box)
labeled_PCA <- plotPCA(vsd_overall, intgroup = c("Treatment"))
labeled_PCA + geom_text_repel(aes_string(x = "PC1", y = "PC2", label = "name"), color = "black", size = 2)

# PCA with different shapes 
new_PCA <- plotPCA(vsd_overall, intgroup = c("Experiment", "Treatment"), returnData = TRUE)
percentVar <- round(100 * attr(new_PCA, "percentVar"))

ggplot(new_PCA, aes(PC1, PC2, color=cell_type, shape=condition)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) + 
  ylab(paste0("PC2: ", percentVar[2], "% variance")) + 
  coord_fixed()
```

Extract results 
```{r}
sub_results <- results(ddsoverall, contrast = c("Treatment","CTR", "IL4IL13"),
                       alpha = 0.5)
#head(sub_results)
sub_results_LFC <- results(ddsoverall,
                       name = "Treatment_IL4IL13_vs_CTR",
                       alpha = 0.5)
```

LFC, MA plot 
```{r}
#plot simple MA plot
plotMA(sub_results_LFC, ylim = c(-10, 8)) #edit
DESeq2::plotMA(sub_results, ylim = c(-10, 10))


# script below not used for extra MAplot...
# Extract the relevant information from DESeqResults object
result_data <- data.frame(
  log2FoldChange = sub_results_LFC$log2FoldChange,
  baseMean = sub_results_LFC$baseMean,
  sig = sub_results_LFC$padj < 0.05  # Creating a binary indicator for significance
)

# Create the plot using ggplot
ggplot(result_data, aes(x = log10(baseMean), y = log2FoldChange, color = sig)) + 
  geom_point()
```

Obtain significant results table and order by padj
```{r}
sub_res_df <- data.frame(sub_results_LFC)
sub_res_sig <- subset(sub_res_df, padj < 0.05) %>%
  arrange(padj)

#Save results table
write.csv(sub_res_sig, file = "./Results IL4_IL13/LogFoldChanges-sig_IL4IL13.csv", row.names = TRUE)

#Save results table
write.csv(sub_res_df, file = "./Results IL4_IL13/LogFoldChanges_full_IL4IL13.csv", row.names = TRUE)
```

Extract the normalized counts
```{r}
normalized_counts <- counts(ddsoverall, normalized = TRUE) #extracted counts normalized by estimateSIzeFactors()
##write.csv(normalized_counts, file.path("./Results IL4_IL13-noSN20/Normalized Counts IL4IL13-noSN20.csv"))
```

plot volcanoplot using enhanced volcano
```{r}
EnhancedVolcano(sub_res_sig,
                lab = rownames(sub_res_sig),
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-10, 13), #edit
                ylim = c(0, 50), #edit
                title = "Effect of IL4/IL13 on hPSC-derived sensory neurons",
                pCutoff = 0.05,
                FCcutoff = 1.5, #1
                pointSize = 2.0,
                labSize = 4.0,
                col=c('grey', 'green4', 'blue', 'red3'),
                colAlpha = 1)
```


Heatmap of all genes 
```{r}
sig_norm_counts <- normalized_counts[rownames(sub_res_sig),]
#display.brewer.all()
heat.colors <- brewer.pal (n = 6, name = "RdYlBu")
#display.brewer.all()

pheatmap(sig_norm_counts, 
         color = colorRampPalette(c("blue", "white", "red"))(50), 
         cluster_rows = T,
         show_rownames = F, 
         annotation = select(metadata, Treatment),
         scale = "row",
         clustering_distance_rows = "euclidean") 
```

```{r}
# Select top genes based on adjusted p-values
top_genes <- rownames(sub_res_sig)[1:50]  # Get the gene names or rownames

# Extract normalized counts for the top significant genes
top_norm_counts <- normalized_counts[top_genes, ]

# Optionally, log-transform the counts (adding 1 to avoid log of 0)
log_transformed_counts <- log2(top_norm_counts + 1)

# Create a heatmap using pheatmap
pheatmap(log_transformed_counts, 
         color = colorRampPalette(c("blue", "white", "red"))(50),
         scale = "row",  # Scale by row to see relative expression
         show_rownames = TRUE,
         cluster_rows = TRUE,
         annotation = select(metadata, Treatment))  # Add condition metadata

```

Heatmap of top50 genes (padj)
```{r}
padj_top50_genes <- head(sub_res_sig, 50)
padj_top50_sig_norm_counts <- merge(normalized_counts, padj_top50_genes, by = 0)
values_padj_top50_sig_norm_counts <- padj_top50_sig_norm_counts[,2:17]

# Set the row names of the dataframe using the "row.names" column
rownames(values_padj_top50_sig_norm_counts) <- padj_top50_sig_norm_counts$Row.names

#padj_top50_genes_ID <- head(coculture_res_ID, 50)
#row.names(values_padj_top50_sig_norm_counts) <- row.names(padj_top50_genes_ID)

pheatmap(values_padj_top50_sig_norm_counts, 
         color = colorRampPalette(c("blue", "white", "red"))(50), 
         cluster_rows = T,
         show_rownames = T, 
         annotation = select(metadata, Treatment),
         scale = "row") 
```

Plot expression of the top20 genes
```{r}
# already have ordered file, just extract top20 genes
sub_top20 <- rownames(sub_res_sig [1:20, ])
# extract normalized count values
sub_top20_norm <- normalized_counts[sub_top20, ]
# use melt to modify the format of the data frame
melted_top20 <- data.frame(melt(sub_top20_norm))
# DID NOT WORK: check the column header in the "melted" data frame
# view(melted_EpCAM_top20)
## add column names that make sense
colnames(melted_top20) <- c("gene", "sample_ID", "normalized_counts")
# add metadata to "melted" dataframe
metadata$sample_ID <- rownames(metadata)
melted_top20 <- merge(melted_top20, metadata)
# plot
ggplot(melted_top20) +
  geom_point(aes(x = gene, y = normalized_counts, color = sample_ID)) +
  scale_y_log10() +
  xlab("Genes") +
  ylab("Normalized Counts") +
  ggtitle("coculture Top 20 Significant DE Genes") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(plot.title=element_text(hjust=0.5)) +
  theme(plot.title.position = "plot")
```

