---
title: "Rscript_DEP_WGCNA"
output: html_document
date: "2023-08-09"
author: "Carli"
update: 
    chunk_output_type: console
---
If necessary, download the required libraries
```{r}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(version = "3.16")

BiocManager::install("DESeq2")
install.packages("dplyr")
install.packages("readxl")
```
Load the data (that is not LOG transformed!)
```{r}
data <- read_excel("cortex_raw.xlsx") #THIS IS THE DATA THAT IS NOT LOG TRANSFORMED
colnames(data) # to see what is the "Genes" column
```

Remove duplicated genes
```{r}
# Are there any duplicated gene names?
data$Genes %>% 
  duplicated() %>% 
  any()

# Make a table of duplicated gene names
data %>% group_by(Genes) %>% summarize(frequency = n()) %>% 
  arrange(desc(frequency)) %>% filter(frequency > 1)

#For further analysis these proteins must get unique names. 

# Make unique names using the annotation in the "Gene.names" column as primary names and the annotation in "Protein.IDs" as name for those that do not have an gene name.
data_unique <- make_unique(data, "Genes", "Protein.Ids")
colnames(data_unique)

# Are there any duplicated names?
#check in the newly created column "name
data_unique$name %>% duplicated() %>% any()
```

#Generate or export a summarized experiment
#First, load an experimental design
experimental_design<- read_excel("metadata_cortex.xlsx")
head(experimental_design)

#Now, generate a se using data_unique and experimental_design
LFQ_columns <- grep("LF", colnames(data_unique))
data_se<-make_se(data_unique, LFQ_columns, experimental_design)

# Let's have a look at the SummarizedExperiment object
data_se

# Plot a barplot of the protein identification overlap between samples
plot_frequency(data_se)

# Filter for proteins that have no missing values (the most stringent filter) IN THIS CASE I WILL NOT USE THIS
#complete_cases <- filter_proteins(data_se, "complete")
#plot_missval(complete_cases) #ERROR: 
#plot_numbers(complete_cases)

# Filter for proteins that are identified in all replicates of at least one condition I DID NOT USE THIS EITHER
#data_filt <- filter_missval(data_se, thr = 0)

# Less stringent filtering:
# Filter for proteins that are identified in 2 out of 3 replicates of at least one condition (I USED THIS)
data_filt2 <- filter_missval(data_se, thr = 1)

# Plot a barplot of the number of identified proteins per samples
plot_numbers(data_filt2)

#data_se is the data before filtering...
plot_numbers(data_se)

# Plot a barplot of the protein identification overlap between samples
#plot_numbers(complete_cases)

#NORMALIZATION

# Normalize the data
data_norm <- normalize_vsn(data_filt2)

meanSdPlot(data_norm)

# Visualize normalization by boxplots for all samples before and after normalization
plot_normalization(data_filt2, data_norm)

# Plot a heatmap of proteins with missing values
plot_missval(data_filt2)
plot_missval(data_norm)

# Plot intensity distributions and cumulative fraction of proteins with and without missing values
plot_detect(data_norm)

#Impute missing data using random draws from a manually defined left-shifted Gaussian distribution (for MNAR)
data_imp_man <- impute(data_norm, fun = "man", shift = 1.5, scale = 0.000001)


# Plot intensity distributions before and after imputation
plot_imputation(data_norm, data_imp_man)

# Visualize normalization by boxplots for all samples before and after normalization
plot_normalization(data_norm, data_imp_man)

#There is no need for imputation in our sampleset because we used the most stringent filtering

write.table(assay(data_imp_man), file="normalized_and_imputed_cortex_nr.csv")
normalized_imp_cortex_nr<-assay(data_imp_man)




##Differential enrichment analysis  based on linear models and empherical Bayes statistics

# Test manually defined comparisons (all comparisons)
data_diff_all_contrasts <- test_diff(data_imp_man, type = "all")
dep <- add_rejections(data_diff_all_contrasts, alpha =0.5, lfc= 0.5) #THIS SET THE THRESHOLDS FOR ADJUSTED P VALUE OF 0.5 AND LOG2FC OF 1

# Plot the first and second principal components
plot_pca(dep, x = 1, y = 2, n = 100, point_size = 4, indicate = c("condition"), label = FALSE)


#Plot the Pearson correlation matrix
plot_cor(dep, significant=TRUE, lower = 0.8, upper = 1, pal = "Reds")

# Plot a heatmap of all significant proteins with the data centered per protein
plot_heatmap(dep,type="centered", 
             clustering_distance="pearson",
             col_limit=3, indicate = "condition")

# Plot a heatmap of all significant proteins (rows) and the tested contrasts (columns)
plot_heatmap(dep, type = "contrast", kmeans = TRUE, col_limit = 4, show_row_names = TRUE)

#Plot a volcano plot for the contrast "WT.KO.Strain_vs_Polb.HT"
plot_volcano(dep, contrast = "WT.KO.Strain_vs_Polb.HT", label_size = 2, add_names = TRUE)

#Plot a volcano plot for the contrast "Polb.HT_vs_WT.C567.Strain"
plot_volcano(dep, contrast = "Polb.HT_vs_WT.C567.Strain", label_size = 2, add_names = TRUE)

#Plot a volcano plot for the contrast "WT.KO.Strain_vs_WT.C567.Strain"
plot_volcano(dep, contrast = "WT.KO.Strain_vs_WT.C567.Strain", label_size = 2, add_names = TRUE)

#Plot a volcano plot for the contrast "Polb.HT_vs_X3xPB_Control"
plot_volcano(dep, contrast = "Polb.HT_vs_X3xPB_Control", label_size = 2, add_names = TRUE)

#Plot a volcano plot for the contrast "Polb.HT_vs_X3xPB_NR"
plot_volcano(dep, contrast = "Polb.HT_vs_X3xPB_NR", label_size = 2, add_names = TRUE)

#Plot a volcano plot for the contrast "X3xPB_Control_vs_X3xPB_NR"
plot_volcano(dep, contrast = "X3xPB_Control_vs_X3xPB_NR", label_size = 2, add_names = TRUE)


# Plot a barplot 
plot_single(dep, proteins = c("Mtnd1"), type="centered", plot=TRUE)

# Extract the results
# Generate a results table
data_results_cortex_DEP <- get_results(dep)

# Number of significant proteins
data_results_cortex_DEP %>% filter(significant) %>% nrow()
# Column names of the results table
colnames(data_results_cortex_DEP)
write.csv(data_results_cortex_DEP, file="data_DEP_cortex.csv")

# Generate a wide data.frame
df_wide <- get_df_wide(dep)
# Generate a long data.frame
df_long <- get_df_long(dep)

#EXPORT THE DATAFRAME OF THIS DEP ANALYSIS
#write.table(df_wide, file="DEP_RESULTS_STRINGENT FILTERING.xls")

##Now we can start with WGCNA PIPELINE
#If not installed yet, please install WCGNA by removing the hashtag from next line
#BiocManager::install("WGCNA", force = TRUE, lib="/Library/Frameworks/R.framework/Versions/4.2/Resources/library")
#library(WGCNA, lib.loc="/Library/Frameworks/R.framework/Versions/4.2/Resources/library")
library(WGCNA)
# The following setting is important, do not omit.
options(stringsAsFactors = FALSE);

#QC and outlier detection
#read normalized data from DEP
proteomic_data <- normalized_imp_cortex_nr

gsg<-goodSamplesGenes(t(proteomic_data))
summary(gsg) #IF THEY ARE ANY GENES THAT ARE DETECTED TO BE OUTLIERS WILL BE SHOWN HERE
gsg$allOK #if this is true, this data passed the test and you don't need to extract those outliers

# if this is false, this data needs to be filtered
#remove genes that are detected as outliers

proteomic_data<-proteomic_data[gsg$goodGenes == TRUE,]


#ANOTHER METHOD TO DETECT OUTLIERS IS TO PERFORM HIERARCHICAL CLUSTERIN
htree<-hclust(dist(t(proteomic_data)), method="average")
plot(htree)
head(htree)
#based on this plot, there is no strong segregation between samples enough to remove one

#based on this plot, only WT.C567.Strain_4 is a clear outlier#Another method to perform quality control is to perform principal component analysis
pca<-prcomp(t(proteomic_data))
pca.dat<-pca$x
pca.var<-pca$sdev^2
pca.var.percent<-round(pca.var/sum(pca.var)*100, digits=2)
library(ggplot2)
pca.dat<-as.data.frame(pca.dat)

ggplot(pca.dat, aes(PC1, PC2))+
         geom_point()+
         geom_text(label = rownames(pca.dat))+
         labs(x=paste0('PC1:', pca.var.percent[1], "%"),
              y=paste0("PC2:", pca.var.percent[2], "%"))
#I decided to remove the sample at the RIGHT
  #based on these plots, only WT.C567.Strain_4 is a clear outlier
#exclude the outlier
samples.to.be.excluded<-c("WT.C567.Strain_4")
data.subset<-proteomic_data[,!(colnames(proteomic_data) %in% samples.to.be.excluded)]

#Network construction
#Choose a set of soft-thresholding powers
power<-c(c(1:10), seq(from=12, to=50, by=2))

#Call the network topology analysis function
sft<-pickSoftThreshold(t(data.subset), #the data.subset was to be transposed!!
                       powerVector = power, 
                       networkType = "signed",
                       verbose=5)

sftdata<-sft$fitIndices
#we will use the max R square values and the minimum mean connectivity
names(sftdata)
#visualize to pick the indices
a1<-ggplot(sftdata,aes(Power, SFT.R.sq, label= Power))+
  geom_point()+
  geom_text(nudge_y=0.1)+
  geom_hline(yintercept=0.8, color="red")+
  labs(x="Power", y="Scale free topology model fit, signed R^2")+
  theme_classic()

#plot for mean connectivity
a2 <- ggplot(sftdata, aes(Power, mean.k., label = Power)) +
  geom_point() +
  geom_text(nudge_y = 0.1) +
  labs(x = 'Power', y = 'Mean Connectivity') +
  theme_classic()

library(gridExtra)
library(grid)
library(ggplot2)
library(lattice)

grid.arrange(a1, a2, nrow = 2)

#Convert matrix to numeric
norm.data<-sapply(proteomic_data, as.numeric)

soft_power<-8
tem_cor<-cor
cor<-WGCNA::cor

#memory estimate blocksize
bwnet<-blockwiseModules(t(data.subset),
                        maxBlockSize = 8000,
                        TOMType = "signed",
                        power = soft_power,
                        mergeCutHeight = 0.25,
                        numericLabels = FALSE,
                        randomseed=1234,
                        verbose=3)


#Module eigengenes
module_eigengenes<-bwnet$MEs

#Print out a preview 
head(module_eigengenes)

table(bwnet$colors)

length(bwnet$dendrograms[[1]]$order)
length(bwnet$colors)

plotDendroAndColors(bwnet$dendrograms[[1]], 
                    cbind(bwnet$unmergedColors, bwnet$colors), 
                    c("unmerged", "merged"), 
                    dendroLabels = FALSE, 
                    addGuide = TRUE,
                    hang= 0.03, 
                    guideHang = 0.05)


# get number of genes for each module
table(bwnet$colors)


#Next step is to use the metadata file to categorize variables to separate samples and to finally correlate modules to traits (metadata variables)

#load metadata file
library(readxl)
library(dplyr)
colData <- read_excel("metadata_cortex.xlsx")

#exclude LFQ.intensity.Sple-14 from the metadata
samples.to.be.excluded<-c("LFQ.intensity.Sple-14")
colData.subset<-colData[!(colData$label %in% samples.to.be.excluded),]

#create traits file (binarize categorical variables)
#traits<-colData%>%
#  mutate(disease_state_bin= ifelse(grepl("AD", disease_state), 1, 0))%>% #this will assign a 1 for AD and a 0 for Control and Control + from both brain areas
#  select(9)

#traits_tau<-colData %>%
#  mutate(tau_bin= ifelse(grepl("Pos", tau_load), 1, 0))%>%
#  select(9)#this will assign a 1 for tau+ samples

#traits_amyloid<-colData%>%
#  mutate(amyloid_bin= ifelse(grepl("Pos", amyloid_load), 1, 0))%>%
#  select(9)#this will assign a 1 for tau+ samples

#To compare everything against everything
colData.subset$condition<- factor(colData.subset$condition, levels = c("WT/KO Strain", "Polb HT", "WT.C567 Strain", "3xPB_Control", "3xPB_NR"))

traits_nr<-binarizeCategoricalColumns(colData.subset$`Treatment Group`,
                                          includePairwise=TRUE,
                                          includeLevelVsAll = FALSE,
                                          minCount = 1)

traits_atp<-c(colData.subset$ATP)
traits_ROS_O2<-c(colData.subset$`ROS 02`)
traits_ROS_H202<-c(colData.subset$`ROS H202`)
traits_RNS_NO<-c(colData.subset$`RNS NO`)
traits_RNS_ONOO<-c(colData.subset$`RNS ONOO-`)
traits_RSS_H2S<-c(colData.subset$`RSS H2S`)
traits_RSS_H2Sn<-c(colData.subset$`RSS H2Sn`)
traits_RCS<-c(colData.subset$`RCS (NBD-H)`)


colnames(colData.subset)


#combining traits data with all the binarized variables
traits<-cbind(traits_nr, 
              traits_atp,
              traits_ROS_O2,
              traits_ROS_H202,
              traits_RNS_NO,
              traits_RNS_ONOO,
              traits_RSS_H2S,
              traits_RSS_H2Sn,
              traits_RCS)
traits1<-traits[,-3]
traits2<-traits1[,-1]

colnames(traits2)

#define number of samples and genes
nSamples<-ncol(data.subset)
nrow(colData.subset)
nGenes<-nrow(proteomic_data)

##Correlation of traits and variables
#correlation of modules and trits using pearson correlation

module_trait_corr<-cor(module_eigengenes, traits2, use="p") #this runs because there are 29 and 29 
module_trait_corr_pvals<-corPvalueStudent(module_trait_corr, nSamples)

#Visualize module trait association as a heatmap

heatmap.data<-merge(module_eigengenes, traits, by="row.names")
head(heatmap.data)

rownames(heatmap.data)<-heatmap.data$Row.names
heatmap.data1<-heatmap.data[-1]



#install.packages("corrplot")
library("corrplot")

corrplot(t(as.matrix(module_trait_corr)),
         method="color",
         p.mat = t((as.matrix(module_trait_corr_pvals))),
         insig = "label_sig",
         tl.col= "black")


#BOX PLOT FOR PLOT OF INTEREST
modules<-tibble::rownames_to_column(module_eigengenes,"sample")
modules$sample<-colData.subset$condition
modules$sample<-as.factor(modules$condition)
modules_joined<-dplyr::inner_join(modules, colData.subset %>%
                                    dplyr::select(`sample label`, condition),
                                  by= c("sample"="condition")
)


p1<- ggplot(modules, aes(x=sample, y=MEskyblue, color=sample)) + 
  geom_boxplot() + theme_test() + theme(legend.position = "none")

p2<-ggplot(modules, aes(x=sample, y=MElightgreen, color=sample)) + 
  geom_boxplot() + theme_test() + theme(legend.position = "none")

p3<-ggplot(modules, aes(x=sample, y=MEyellow, color=sample)) + 
  geom_boxplot() + theme_test() + theme(legend.position = "none")

p4<-ggplot(modules, aes(x=sample, y=MEsalmon, color=sample)) + 
  geom_boxplot() + theme_test() + theme(legend.position = "none")

p5<-ggplot(modules, aes(x=sample, y=MEpaleturquoise, color=sample)) + 
  geom_boxplot() + theme_test() + theme(legend.position = "none")

p6<-ggplot(modules, aes(x=sample, y=MEcyan, color=sample)) + 
  geom_boxplot() + theme_test() + theme(legend.position = "none")



grid.arrange(p1) 



#Get the genes/proteins that are included in the modules for amyloid association
module.gene.mapping<-as.data.frame(bwnet$colors)
module.gene.mapping%>%
  filter(`bwnet$colors`==c("blue")%>% #i use royalblue becuase it was positively associated to tau load in the WGCNA 
  rownames()

#in order to get a list and put it into metascape or do downstream Gene Ontology
amyloid_associated_module<-module.gene.mapping%>% 
  filter(`bwnet$`==c("blue"))%>%
  rownames()


#to get all modules
write.table(module.gene.mapping, file="module_gene_mapping.csv")




